---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Yaasiel Leos Ledesma"
date: "2026-02-27"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

[deployed html url](https://yaasiell.github.io/workshop-08_winter-2026/code/workshop-code.html)

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschscholzia californica_   | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  path = here(
    "data",
    "Valliere_etal_EcoApps_Data.xlsx"
  ),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw()) # set aesth. for all plots you're making
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# creating a new object called 'drought_exp_clean'
drought_exp_clean <- drought_exp |>
  # cleaning up column names
  clean_names() |>
  # creating a new column 
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |>
  # changes water treatment column names to corresponding full names 
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  )) |>
  # 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |>
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |>
  # selecting columns of interest. 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: checking strucutre
str(drought_exp_clean)
```



Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: displaying rows in data frame

slice_sample(
  drought_exp_clean,
  n = 10
)
```




# 3. Visualizing data

## What is the relationship between SLA and plant mass?  


Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______).

```{r}
#| label: sla and mass plot
 # base layer : ggplot
  ggplot(data = drought_exp_clean,
         aes(x = sla,
             y = total_g)) +
  geom_point(color = "#50a492") +
  # labeling x and y axis
  labs(x = "Specific Lead Area (cm\U00B2/g)",
       y = "Total Mass (g)",
       title = "As SLA increases, total mass increases")
```

 





## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.



```{r}
#| label: water treament and mass plot

ggplot(data = drought_exp_clean,
       aes(x = water_treatment,
           y = total_g,
           color = water_treatment)) +
  geom_jitter(height = 0,
              width = 0.2) +
  scale_color_manual(values = c(
    "Drought stressed" = ds_col,
    "Well watered" = ww_col
  )) +
  labs(x = "Water treatment",
       y = "Total mass (g)",
       title = "On average, plant mass in well watered treatment\n is higher than plant mass in drought stressd treamtment")
```




## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass _and_ wrap the long species names using the `label_wrap()` function from the `scales` package.


```{r}
#| label: species and plant mass plot

# base layer: ggplot
ggplot(data = drought_exp_clean,
       aes(x = species_name,
           y = total_g,
           color = species_name)) +
  geom_jitter(height = 0,
              width = 0.2) +
  scale_color_manual(values = c(
    "Lotus scoparius" = lotsco_col,
    "Penstemon centranthifolius" = pencen_col,
    "Salvia leucophylla" = salleu_col,
    "Encelia californica" = enccal_col,
    "Stipa pulchra" = stipul_col,
    "Eschscholzia californica" = esccal_col,
    "Grindelia camporum" = gricam_col
  )) +
  scale_x_discrete(labels = label_wrap(10)) +
  labs(x = "Species Name",
       y = "Total mass(g)",
       title = "Plant mass differs across species") +
  theme(legend.position ="none")
```






# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1, # no predictors in this model
  data = drought_exp_clean
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name,
  data = drought_exp_clean
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment,
  data = drought_exp_clean
)
# model 3: SLA and species

model3 <- lm(
  total_g ~ sla + species_name,
  data = drought_exp_clean
)
# model 4: water treatment and species

model4 <- lm(
  total_g ~ water_treatment + species_name,
  data = drought_exp_clean
)
# model 5: SLA
model5 <- lm(
  total_g ~ sla,
  data = drought_exp_clean
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment,
  data = drought_exp_clean
)

# model 7: species
model7 <- lm(
  total_g ~ species_name,
  data = drought_exp_clean
)
```

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#| label: model selection

AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  arrange(AICc)
```

The best model to describe variation in plant mass only includes water treatment and species. 

## Model diagnostics

Insert code to look at the diagnostic plots for the best model.


```{r}
#| label: checking diagnostics

par(mfrow = c(2,2)) # display plots in a 2x2 grid
plot(model4) #model to check diagnostics

```
Diagnostics plot look good 
Residuals (standing in for errors) 
No outliers influencing model estimates of coefficients.

## Summary

Insert code to display the summary for the best model. 
```{r}
#| label: looking at model summary

summary(model4)
```



**What are the reference levels?**  

Drought stressed and Lotus scoparius. 

**What does the `(Intercept)` represent?**  

The intercept is the model prediction for plant mass for Lotus scoparius in the drought stressed treatment.The model predicts that lotus scoparius mass in the drought stressed treatment will be 0.055 $\pm$ 0.025 (SE) g. 

**What does `water_treatmentWell watered` represent?**

This is the model prediction for the difference for Lotus scoparius mass between the well watered and the drought stressed treatments. The model predicts that this difference is 0.12 $\pm$ 0.017 g. 

**What does `species_namePenstemon centranthifolius` represent?**  
This is the model prediction for the difference between Penstemon centranthifolius mass in the well watered treatment and Lotus scoparius mass in the well watered treatment. The model predicts that this differnce is 0.05 $\pm$ 0.03 g. 



**What does `species_nameEschscholzia californica` represent?**

This is the model prediction for the diffrence between Eschscholzia californica mass in the well watered treatment and Lotus scoparius mass in the well watered treatment. The model predicts that this difference is 0.23 $\pm$ 0.03g. 


## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

72% of the cariation in plant mass can be explained by this model.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Treatment influences plant mass.

Species influences plant mass. 

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: genrate predictions

preds <- ggpredict(
  model4,
  terms = c("water_treatment","species_name")
) |>
  rename(water_treatment = x,
         species_name = group)
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.
```{r}
ggplot(data = drought_exp_clean,
       aes(x = species_name,
           y = total_g,
           color = water_treatment)) +
  geom_jitter(height = 0,
              width = 0.2,
              alpha = 0.5) +
  geom_pointrange(data = preds,
                  aes(x = water_treatment,
                      y = predicted,
                      ymin = conf.low,
                      ymax = conf.high)) +

  scale_color_manual(values = c(
    "Lotus scoparius" = lotsco_col,
    "Penstemon centranthifolius" = pencen_col,
    "Salvia leucophylla" = salleu_col,
    "Encelia californica" = enccal_col,
    "Stipa pulchra" = stipul_col,
    "Eschscholzia californica" = esccal_col,
    "Grindelia camporum" = gricam_col
  )) +
  scale_x_discrete(labels = label_wrap(10)) +
  labs(x = "Species Name",
       y = "Total mass(g)",
       title = "Plant mass differs across species") +
  theme(legend.position ="none") +
  facet_wrap(~ species_name,
             nrow = 1)
```


# 6. Writing

## Figure caption








## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**test statistic**


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

#### Goal of analysis


#### Model selection


#### Model diagnostics


### Results









